---
title: "distfromq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{distfromq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 8,
  out.width = "100%"
)
```

# Introduction

The `distfromq` package provides functions for constructing an approximation to a probability distribution from a finite set of quantiles of the distribution. The package provides four main functions:

1. `make_p_fn` returns a function that is analogous to `pnorm`, and can be used to evaluate the approximated cumulative distribution function (CDF).
2. `make_d_fn` returns a function that is analogous to `dnorm`, and can be used to evaluate the approximated  probability density function (PDF).
3. `make_q_fn` returns a function that is analogous to `qnorm`, and can be used to evaluate the approximated quantile function (QF), i.e., the inverse CDF.
4. `make_r_fn` returns a function that is analogous to `rnorm`, and can be used to simulate a pseudo-random sample from the approximated distribution.

All four functions take a set of probability levels (`ps`) and the corresponding quantiles (`qs`) of the distribution to be approximated as inputs, as well as arguments specifying the methods for interpolating those quantiles to estimate the interior of the distribution, and for extrapolating into the tails of the distribution using a specified parametric family.

The next section of the vignette describes the methods that are used for approximating the distribution, and the following section gives examples.

# Methods description

Suppose that at the probability levels $\tau_1, \ldots, \tau_K \in (0, 1)$, we have available the corresponding quantiles $q_1 = F^{-1}(\tau_1), \ldots, q_K = F^{-1}(\tau_K)$ for some distribution with CDF $F$. Our goal is to reconstruct an estimate of $F$, the corresponding PDF $f$, and the QF $F^{-1}$, as well ast a method for sampling from the distribution.

Our approach takes different strategies for interpolating on the interior of the provided quantiles and for extrapolating into the tails; we outline these procedures briefly here and give more detail below:

1. We estimate the distribution on the interior of the predictive quantiles using the following procedure:
    * If there are repeated quantiles across different probability levels (i.e., $q_k = q_{k'}$ for some indices $k \neq k'$) indicating the presence of one or more discrete point masses, we will represent the distribution as a mixture of a continuous component and a discrete component. To facilitate that, we split the quantiles into groups corresponding to those two components.
    * For the CDF, we estimate the continuous component using a monotonic spline that interpolates the quantiles, along with discrete point masses at any points where there is a repeated quantile for multiple distinct probability levels.
    * For the QF, we invert the estimate of the CDF.
    * For the PDF, we use the derivative of the spline-based approximation to the CDF if there are no discrete components.
2. we approximate the tails with a distribution in a specified location-scale family.

In cases where there are repeated quantiles, we take the discrete component of the distribution to have a point mass at each quantile value that is repeated. The discrete distribution assigns probability to each such point $q$ corresponding to $\max(\{\tau_{k}: q_k = q\}) - \min(\{\tau_{k}: q_k = q\})$. The continuous distribution is then estimated based on a set of adjusted quantile pairs that is obtained by eliminating duplicate quantiles and subtracting the discrete distribution.

To approximate the CDF of the continuous distribution component at interior points, we fit a monotonic cubic spline that interpolates the set of "observations" $\{(q_1, \tau_1), \ldots, (q_K, \tau_K)\}$ (where these now denote the adjusted quantiles for the continuous distribution component). This spline is an estimate of the CDF. Because the spline passes through the points $(q_1, \tau_1)$ and $(q_K, \tau_K)$, the integral of its derivative over the interval $[q_1, q_K]$ is equal to $\tau_K - \tau_1$.

To approximate the QF, we invert the CDF estimate. For the continuous component, in principle the inverse can be calculated analytically by directly inverting the cubic spline. However, we have noticed that this can yield a QF estimate that is not a precise inverse of the CDF estimate due to numerical instabilities in the inversion. Therefore, as an alternative we allow for the use of a piecewise linear approximation to the spline that is obtained by evaluating the cdf at a grid of quantile values. The inverse of this piecewise linear function can be calculated with more numerical stability, and this approach is recommended when it is critical that the CDF and QF be inverses.

We estimate the distribution for the left and right tails separately, assuming that they come from a specified location-scale family which may be different in each tail. Setting notation, suppose that $Y = a + b \cdot Z$ where the random variable $Z$ has a specified distribution. Recall that at the probability level $\tau$, a quantile of $Y$ can be calculated in terms of the corresponding quantile of $Z$ via $F_Y^{-1}(\tau) = a + b \cdot F_Z^{-1}(\tau)$. Using the quantiles at two probability levels $\tau_i$ and $\tau_j$, we can calculate the value of $b$ using

\begin{align*}
\frac{F_Y^{-1}(\tau_i) - F_Y^{-1}(\tau_j)}{F_Z^{-1}(\tau_i) - F_Z^{-1}(\tau_j)} &= \frac{a + b \cdot F_Z^{-1}(\tau_i) - (a + b \cdot F_Z^{-1}(\tau_j))} {F_Z^{-1}(\tau_i) - F_Z^{-1}(\tau_j)} \\
&= b \cdot \frac{F_Z^{-1}(\tau_i) - F_Z^{-1}(\tau_j)}{F_Z^{-1}(\tau_i) - F_Z^{-1}(\tau_j)} \\
&= b
\end{align*}

Similarly, we can calculate the value of $a$ as

$$F_Y^{-1}(\tau_i) - b \cdot F_Z^{-1}(\tau_i) = a + b \cdot F_Z^{-1}(\tau_i) - b \cdot F_Z^{-1}(\tau_i) = a$$

In the above expressions, we use the two smallest quantiles when estimating the lower tail and the two largest quantiles when estimating the upper tail. With these choices, by construction the lower tail integrates to $\tau_1$ on the interval $(-\infty, q_1]$ and the upper tail integrates to $1 - \tau_K$ on the interval $[q_K, \infty)$.

As an alternative, we also allow for the use of a log-normal distribution for extrapolation into the tails. The parameters of the log-normal distribution are estimated by applying the approach outlined above to the logarithms of the quantiles.

Random deviates are obtained by sampling $U \sim \text{Unif}(0,1)$ and then evaluating the approximated quantile function: $Y = F_Y^{-1}(U)$.

# Illustration of package functionality

We illustrate the package functionality with a series of three examples. The first example is a setting where the distribution is for a continuous random variable with a defined probability density funciton. The second and third examples illustrate the behavior of the functions in settings where the provided quantiles are the same at multiple probability levels, indicating that there is some discrete component to the distribution.

```{r setup}
library(distfromq)
library(ggplot2)
library(dplyr)
```

## Example 1

In our first example, we work with quantiles from the distribution $Y \sim \text{log normal}(4, 0.5)$. We compare this predictive distribution with three approximations of it that are derived from the quantiles using different assumptions for the tail behavior:

1. The first assumes a log-normal distribution for the tails. This method should reconstruct the true distribution.
2. The second assumes a normal distribution for the tails. This method has a mismatch in the support of the lower tail (including values less than 0), and has a thinner upper tail than the log-normal.
3. The third assumes a Cauchy distribution for the tails. This method has a mismatch in the support of the lower tail (including values less than 0), and has a heavier upper tail than the log-normal.

```{r}
quantile_probs <- seq(from = 0.1, to = 0.9, by = 0.1)

meanlog <- 4.0
sdlog <- 0.5
q_lognormal <- qlnorm(quantile_probs, meanlog = meanlog, sdlog = sdlog)
```

```{r}
x <- seq(from = 0.0, to = 400.0, length = 501)
cdf_lognormal <- plnorm(x, meanlog = meanlog, sdlog = sdlog)

p_lognormal_approx <- make_p_fn(ps = quantile_probs,
                                qs = q_lognormal,
                                lower_tail_dist = "lnorm",
                                upper_tail_dist = "lnorm")
cdf_lognormal_approx <- p_lognormal_approx(x)


p_normal_approx <- make_p_fn(ps = quantile_probs,
                             qs = q_lognormal,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm")
cdf_normal_approx <- p_normal_approx(x)

p_cauchy_approx <- make_p_fn(ps = quantile_probs,
                             qs = q_lognormal,
                             lower_tail_dist = "cauchy",
                             upper_tail_dist = "cauchy")
cdf_cauchy_approx <- p_cauchy_approx(x)

dplyr::bind_rows(
    data.frame(
        x = x,
        y = cdf_lognormal,
        dist = "Log normal"
    ),
    data.frame(
        x = x,
        y = cdf_lognormal_approx,
        dist = "Spline interpolation,\nlog-normal tails"
    ),
    data.frame(
        x = x,
        y = cdf_normal_approx,
        dist = "Spline interpolation,\nnormal tails"
    ),
    data.frame(
        x = x,
        y = cdf_cauchy_approx,
        dist = "Spline interpolation,\nCauchy tails"
    )
) %>%
    ggplot() +
        geom_line(
            mapping = aes(x = x, y = y, color = dist, linetype = dist),
            size = 0.8) +
        geom_point(
            data = data.frame(q = q_lognormal, p = quantile_probs),
            mapping = aes(x = q, y = p),
            size = 1.2
        ) +
        scale_color_viridis_d(
            "Distribution",
            end = 0.9
        ) +
        scale_linetype_discrete("Distribution") +
        ylab("Probability") +
        xlab("") +
        theme_bw()
```

```{r}
d_lognormal_approx <- make_d_fn(ps = quantile_probs,
                                qs = q_lognormal,
                                lower_tail_dist = "lnorm",
                                upper_tail_dist = "lnorm")

d_normal_approx <- make_d_fn(ps = quantile_probs,
                             qs = q_lognormal,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm")

d_cauchy_approx <- make_d_fn(ps = quantile_probs,
                             qs = q_lognormal,
                             lower_tail_dist = "cauchy",
                             upper_tail_dist = "cauchy")

pdf_lognormal <- dlnorm(x, meanlog = meanlog, sdlog = sdlog)
pdf_lognormal_approx <- d_lognormal_approx(x)
pdf_normal_approx <- d_normal_approx(x)
pdf_cauchy_approx <- d_cauchy_approx(x)

dplyr::bind_rows(
    data.frame(
        x = x,
        y = pdf_lognormal,
        dist = "Log normal"
    ),
    data.frame(
        x = x,
        y = pdf_lognormal_approx,
        dist = "Spline interpolation,\nlog-normal tails"
    ),
    data.frame(
        x = x,
        y = pdf_normal_approx,
        dist = "Spline interpolation,\nnormal tails"
    ),
    data.frame(
        x = x,
        y = pdf_cauchy_approx,
        dist = "Spline interpolation,\nCauchy tails"
    )
) %>%
    ggplot() +
        geom_vline(
            data = data.frame(q = q_lognormal),
            mapping = aes(xintercept = q),
            size = 0.2
        ) +
        geom_line(
            mapping = aes(x = x, y = y, color = dist, linetype = dist),
            size = 0.8) +
        scale_color_viridis_d(
            "Distribution",
            end = 0.9
        ) +
        scale_linetype_discrete("Distribution") +
        ylab("Probability Density") +
        xlab("") +
        theme_bw()
```

```{r}
r_normal_approx <- make_r_fn(ps = quantile_probs,
                             qs = q_lognormal,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm")
r_lognormal_approx <- make_r_fn(ps = quantile_probs,
                                qs = q_lognormal,
                                lower_tail_dist = "lnorm",
                                upper_tail_dist = "lnorm")
r_cauchy_approx <- make_r_fn(ps = quantile_probs,
                             qs = q_lognormal,
                             lower_tail_dist = "cauchy",
                             upper_tail_dist = "cauchy")
                             
normal_approx_sample <- r_normal_approx(n=10000)
lognormal_approx_sample <- r_lognormal_approx(n=10000)
cauchy_approx_sample <- r_cauchy_approx(n=10000)

bind_rows(
    data.frame(x=normal_approx_sample, dist = "Spline interpolation,\nnormal tails"),
    data.frame(x=lognormal_approx_sample, dist = "Spline interpolation,\nlog-normal tails"),
    data.frame(x=cauchy_approx_sample, dist = "Spline interpolation,\nCauchy tails")
) %>%
    ggplot() +
        geom_density(mapping = aes(x = x, color = dist, linetype = dist)) +
        scale_color_viridis_d(
            "Distribution",
            end = 0.9
        ) +
        scale_linetype_discrete("Distribution") +
        theme_bw()
```


```{r}
bind_rows(
    data.frame(x=normal_approx_sample, dist = "Spline interpolation,\nnormal tails"),
    data.frame(x=lognormal_approx_sample, dist = "Spline interpolation,\nlog-normal tails"),
    data.frame(x=cauchy_approx_sample, dist = "Spline interpolation,\nCauchy tails")
) %>%
    ggplot() +
        geom_density(mapping = aes(x = x, color = dist, linetype = dist)) +
        scale_color_viridis_d(
            "Distribution",
            end = 0.9
        ) +
        scale_linetype_discrete("Distribution") +
        xlim(-100, 300) +
        theme_bw()
```

```{r}
x <- seq(from = 0.0, to = 400.0, length = 501)
```

```{r}
ps <- seq(from = 0.01, to = 0.99, by = 0.01)

q_normal_approx <- make_q_fn(ps = quantile_probs,
                             qs = q_lognormal,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm")
q_lognormal_approx <- make_q_fn(ps = quantile_probs,
                                qs = q_lognormal,
                                lower_tail_dist = "lnorm",
                                upper_tail_dist = "lnorm")
q_cauchy_approx <- make_q_fn(ps = quantile_probs,
                             qs = q_lognormal,
                             lower_tail_dist = "cauchy",
                             upper_tail_dist = "cauchy")

quantiles_lognormal <- qlnorm(ps, meanlog = meanlog, sdlog = sdlog)
quantiles_normal_approx <- q_normal_approx(ps)
quantiles_lognormal_approx <- q_lognormal_approx(ps)
quantiles_cauchy_approx <- q_cauchy_approx(ps)


dplyr::bind_rows(
    data.frame(
        x = ps,
        y = quantiles_lognormal,
        dist = "Log normal"
    ),
    data.frame(
        x = ps,
        y = quantiles_normal_approx,
        dist = "Spline interpolation,\nnormal tails"
    ),
    data.frame(
        x = ps,
        y = quantiles_lognormal_approx,
        dist = "Spline interpolation,\nlognormal tails"
    ),
    data.frame(
        x = ps,
        y = quantiles_cauchy_approx,
        dist = "Spline interpolation,\nCauchy tails"
    )
) %>%
    ggplot() +
        geom_line(
            mapping = aes(x = x, y = y, color = dist, linetype = dist),
            size = 0.8) +
        scale_color_viridis_d(
            "Distribution",
            end = 0.9
        ) +
        scale_linetype_discrete("Distribution") +
        ylab("Quantile") +
        xlab("Probability Level") +
        theme_bw()
```

## Example 2: duplicated quantiles on the interior

When quantile values are the same at multiple probability levels, the distribution assigns non-zero probability to that value, corresponding to a "step" or "jump" in the cumulative distribution function. We note that in this case, `make_d_fn` will throw an error because a continuous density does not exist. We illustrate the behavior of the estimates provided by `make_p_fn` and `make_q_fn` in the presence of repeated quantiles in this and the following example.

```{r, error=TRUE}
quantile_probs <- seq(from = 0.1, to = 0.9, by = 0.1)
quantile_values <- c(1.0, 2.0, 3.0, 3.0, 3.0, 3.0, 3.0, 8.0, 9.0)

d_normal_approx <- make_d_fn(ps = quantile_probs,
                             qs = quantile_values,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm")
p_normal_approx <- make_p_fn(ps = quantile_probs,
                             qs = quantile_values,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm")
q_normal_approx <- make_q_fn(ps = quantile_probs,
                             qs = quantile_values,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm")
r_normal_approx <- make_r_fn(ps = quantile_probs,
                             qs = quantile_values,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm")


x <- seq(from = 0.0, to = 20.0, length = 5001)
cdf_normal_approx <- p_normal_approx(x)
ggplot() +
  geom_line(data=data.frame(x=x, y=cdf_normal_approx),
            mapping = aes(x=x, y=y)) +
  geom_point(data=data.frame(x=quantile_values, y=quantile_probs),
             mapping = aes(x=x, y=y))

ps <- seq(from = 0.0, to = 1.0, length = 5001)
qf_normal_approx <- q_normal_approx(ps)
ggplot() +
  geom_line(data=data.frame(p=ps, y=qf_normal_approx),
            mapping = aes(x=p, y=y)) +
  geom_point(data=data.frame(x=quantile_probs, y = quantile_values),
             mapping = aes(x=x, y=y))

samples_normal_approx <- r_normal_approx(n = 10000)
mean(samples_normal_approx == 3.0)
```

As described earlier, in general the estimated cdf and quantile function are not precise inverses due to numerical issues that arise when inverting the monotonic spline. We illustrate below that the estimated cdf composed with the estimated quantile function may not be an identity function, even in regions where the distribution is continuous.

```{r}
ps <- seq(from = 0.0, to = 1.0, length = 101)
out_ps <- p_normal_approx(q_normal_approx(ps))
out_ps

ps[(ps < 0.3) | (ps > 0.7)] - out_ps[(ps < 0.3) | (ps > 0.7)]
```

These errors can be mitigated by using a piecewise linear representation of the cdf estimate:

```{r}
p_normal_approx_lin <- make_p_fn(ps = quantile_probs,
                             qs = quantile_values,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm",
                             interior_args = list(n_grid = 20))
q_normal_approx_lin <- make_q_fn(ps = quantile_probs,
                             qs = quantile_values,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm",
                             interior_args = list(n_grid = 20))

ps <- seq(from = 0.0, to = 1.0, length = 101)
out_ps <- p_normal_approx_lin(q_normal_approx_lin(ps))
out_ps

ps[(ps < 0.3) | (ps > 0.7)] - out_ps[(ps < 0.3) | (ps > 0.7)]
```

We note that the two representations of the CDF are generally quite similar, but the QF estimates can differ substantially:

```{r}
x <- seq(from = 0.0, to = 20.0, length = 5001)
plot_df <- rbind(
    data.frame(
        x = x,
        cdf = p_normal_approx(x),
        type = "spline"
    ),
    data.frame(
        x = x,
        cdf = p_normal_approx_lin(x),
        type = "piecewise linear"
    )
)
ggplot() +
  geom_line(data = plot_df,
            mapping = aes(x = x, y = cdf, color = type, linetype = type)) +
  geom_point(data = data.frame(x = quantile_values, y = quantile_probs),
             mapping = aes(x = x, y = y))
```

```{r}
ps <- seq(from = 0.0, to = 1.0, length = 5001)
plot_df <- rbind(
    data.frame(
        p = ps,
        qf = q_normal_approx(ps),
        type = "spline"
    ),
    data.frame(
        p = ps,
        qf = q_normal_approx_lin(ps),
        type = "piecewise linear"
    )
)
ggplot() +
    geom_line(data = plot_df,
              mapping = aes(x = p, y = qf, color = type, linetype = type)) +
    geom_point(data = data.frame(x = quantile_probs, y = quantile_values),
               mapping = aes(x = x, y = y))
```

We therefore recommend using the piecewise linear representation when using a quantile function.

## Example 3: duplicated quantiles in the tails

Our final example illustrates the behavior when there are duplicated quantiles at the edges. In these cases, we ignore the specified `lower_tail_dist` and/or `upper_tail_dist`, and assume that the point mass at the edge contains all of the tail probability.

```{r}
quantile_probs <- seq(from = 0.1, to = 0.9, by = 0.1)
quantile_values <- c(1.0, 1.0, 3.0, 3.0, 3.0, 3.0, 3.0, 9.0, 9.0)

p_normal_approx <- make_p_fn(ps = quantile_probs,
                             qs = quantile_values,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm")
p_normal_approx_lin <- make_p_fn(ps = quantile_probs,
                             qs = quantile_values,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm",
                             interior_args = list(n_grid = 20))
q_normal_approx <- make_q_fn(ps = quantile_probs,
                             qs = quantile_values,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm")
q_normal_approx_lin <- make_q_fn(ps = quantile_probs,
                             qs = quantile_values,
                             lower_tail_dist = "norm",
                             upper_tail_dist = "norm",
                             interior_args = list(n_grid = 20))

x <- seq(from = 0.0, to = 20.0, length = 5001)
plot_df <- rbind(
    data.frame(
        x = x,
        cdf = p_normal_approx(x),
        type = "spline"
    ),
    data.frame(
        x = x,
        cdf = p_normal_approx_lin(x),
        type = "piecewise linear"
    )
)
ggplot() +
  geom_line(data = plot_df,
            mapping = aes(x = x, y = cdf, color = type, linetype = type)) +
  geom_point(data = data.frame(x = quantile_values, y = quantile_probs),
             mapping = aes(x = x, y = y))
```


```{r}
ps <- seq(from = 0.0, to = 1.0, length = 5001)
plot_df <- rbind(
    data.frame(
        p = ps,
        qf = q_normal_approx(ps),
        type = "spline"
    ),
    data.frame(
        p = ps,
        qf = q_normal_approx_lin(ps),
        type = "piecewise linear"
    )
)
ggplot() +
    geom_line(data = plot_df,
              mapping = aes(x = p, y = qf, color = type, linetype = type)) +
    geom_point(data = data.frame(x = quantile_probs, y = quantile_values),
               mapping = aes(x = x, y = y))

```
